### 2013-11-08 src2/thesis.tex
update

A CUDA kernel that uses shared memory takes the form:

@@

\FloatBarrier 
\subsection{Writing efficient GPU Kernels} 
 \emph{now takes up both CUDA and Obsidian aspects at the same time, pref. with code examples}

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% GPU Programming 
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------

\subsection{CUDA programming}

A kernel that uses shared memory takes the form:

\subsection{Writing efficient GPU kernels} 
\label{sec:efficient}

Writing efficient GPU kernels is hard. There are many details that the CUDA 
programmer must be aware of in order to make optimal use of the GPU. 
The NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
to learn more about these details. Here is a list, with explanations of some 
of the practices that the guide deems most important.

@@

@@

@@

@@

\item {\bf Avoid different execution paths within a block:} Any branching instruction that 
causes the execution paths to diverge within a warp will affect performance. When the execution 
paths within a warp diverge, the computation is serialised.

\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} Extreme care must be taken 
when synchronising within any conditionally executed code; every thread must reach this barrier.
Failing to ensure this will likely lead to the kernel producing incorrect results. Note that 
this is a major limitation to the composability of CUDA functions. Wheather or not it is safe to 
call some function in a certain place in your code can be decided only by inspecting that function's 
implementation. 
%f it is safe or 
%not to call some function at a certain place in your code, can only be decided by inspecting 
%that functions implementation. 
\end{itemize}

@@

%\citet{FERMIOPT}
%\citet{BestPrac} 
%\citet{merrill}

>>>>

\pagebreak

\noindent A CUDA kernel that uses shared memory takes the form:

@@

This change introduced some noise. The reason for this is that Obsidian programs that 
use memory are monadic. The {\tt BProgram} in the result type is a monad. Another difference 
is that we need to be able to write the array elements to memory; in Obsidian this means 
that the element type must be in the {\tt MemoryOps} class.

\FloatBarrier 
\subsection{Writing efficient GPU Kernels} 
\label{sec:efficient}

Writing efficient GPU kernels is hard. There are many details that the CUDA 
programmer must be aware of in order to make optimal use of the GPU. 
The NVIDIA ``CUDA C Best Practices Guide''~\citet{BestPrac}, is a good source 
to learn more about these details. Here is a list, with explanations of some 
of the practices that the guide deems most important. The list also contains 
information about how these best practices influence our work on Obsidian.

@@

Data transfers between the host memory and device (GPU) memory is out of Obsidian's scope. 
If Obsidian is used to generate kernels for a computation, the programmer has already decided that 
a GPU should be used. However, even after making the decision to use the GPU the programmer 
may need to consider how to transfer the data. In some cases it is possible to overlap 
computation on the GPU and data transfer. These kinds of optimisations are considered 
when writing the C program that launches the CUDA kernels. Obsidian does have rudimentary 
support for launching kernels from within Haskell but currently we consider this mostly a
tool for quickly being able to run a kernel and see what it does, not as a means to implement 
applications. This part of Obsidian is future work.

@@

Inputs to an Obsidian program are pull arrays. Pull arrays can be arbitrarily permuted; 
the permutation used will decide the access pattern. If this method can be applied or 
not partly also depends on the algorithm being implemented.

@@

Intermediate arrays are stored into shared memory when the Obsidian programmer uses 
the {\tt force} function.

@@

Obsidian requires regularity even more than CUDA does. Since we focus on generating high 
performance kernels, we rule out generation of kernels that work on different sizes (per block). 
The choice of how many threads to use per block is in the hands of the Obsidian programmer.
The method for making that choice has changed over time. Earlier, the number of threads 
needed by a kernel was decided by the size of the arrays it computed. This has recently changed
and is now set by a parameter passed to the code generation process.

\item {\bf Avoid different execution paths within a block:} Any branching instruction that 
causes the execution to branch into different code (in CUDA jargon called a diverging branch) 
within a warp will affect performance. When the execution paths within a warp diverge, the 
computation is serialised.

Sometimes divergent code cannot be avoided. However, it is important not to have threads 
diverge unnecessarily. When using pull arrays alone this was often a problem; with the 
addition of push arrays the situation improved. Using push arrays some divergent computations 
can be explicitly split up into phases, each phase with all threads following the 
same execution path. This means that fewer threads are required to do the same work.

\item {\bf Do not use {\tt \_\_synchthreads()} in diverging code:} Extreme care must be taken 
when synchronising within any conditionally executed code; every thread must reach this barrier.
Failing to ensure this will likely lead to the kernel producing incorrect results. Note that 
this is a major limitation to the composability of CUDA functions. Wheather or not it is safe to 
call some function in a certain place in your code can be decided only by inspecting that 
function's implementation.

In Obsidian it is impossible for a barrier synchronisation to appear within diverging 
conditional code. Conditionals can be introduced either by an {\tt ifThenElse} function 
at the element level or by a {\tt Cond} statement that is an operation in the {\tt Program} 
monad. The {\tt Program} monad is parameterised on the GPU hierarchy level. There are 
thread programs ({\tt TProgram}), block programs ({\tt BProgram}) and grid programs ({\tt GProgram}). the {\tt Cond} conditional choose between two thread programs while a synchronisation can 
only appear in an Obsidian block program. More information about programs and the GPU hierarchy 
is available in section~\ref{sec:OBSIMP}.

%f it is safe or 
%not to call some function at a certain place in your code, can only be decided by inspecting 
%that functions implementation. 
\end{itemize}

@@

% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
%
% Obsidian Implementation 
%
% ---------------------------------------------------------------------------
% ---------------------------------------------------------------------------
\subsection{Obsidian Implementation}
\label{sec:OBSIMP}
